### 2024-03-03 paper/paper.tex
feat: set up initial document (incl. settings)

\documentclass[12pt,a4paper,oneside]{article}
%\documentclass[12pt,a4paper,oneside,twocolumn]{article}

% Bibliography & Language
\usepackage[english]{babel} % Language-specific stuff, f.i. hyphenation
\usepackage{csquotes} % Allows for quotes - recommended by babel
\usepackage[style=ieee, sorting=nyt, urldate=long, date=long, backend=biber]{biblatex}
\addbibresource{sources.bib}

% General document
\usepackage[useregional]{datetime2} % Date
\usepackage{titling} % Allows for multiple titles (info sheet)
\usepackage{libertinus} % Fonts
\usepackage[margin=3cm]{geometry} % Custom margins
\usepackage{parskip} % Removes indentation at start of paragraph, add space between them
\usepackage[normalem]{ulem} % Underlining
\useunder{\uline}{\ul}{}
\usepackage[htt]{hyphenat} % Hyphenation for monospace text
\usepackage{todonotes} % Add todos
\usepackage{blindtext} % lorem ipsum, blind text

% Graphics
\usepackage{graphicx} % Graphics extended
\usepackage{hyperref} % Refer to things, including URLs
\usepackage{float} % Allows for H float option ("here")
\usepackage{caption} % allows for centered figure caption justification

% Report hand-in date
\DTMsavedate{date}{2024-03-29}

\title{\Huge{Machine Learning}\\\LARGE TODO: Title goes here}
\author{
Kai E. Niermann \and
Dávid Miklo \and 
Trix Taicet \and
Red Kaláb \and
Conner Dassen
}
\date{\DTMusedate{date}}
\begin{document}
\maketitle

\blinddocument

\include{info_sheet.tex}
\end{document}

### 2024-03-07 paper/paper.tex
basic paper outline and some formatting changes

@@
 %\documentclass[12pt,a4paper,oneside,twocolumn]{article}
-
-% Bibliography & Language
-\usepackage[english]{babel} % Language-specific stuff, f.i. hyphenation
-\usepackage{csquotes} % Allows for quotes - recommended by babel
-\usepackage[style=ieee, sorting=nyt, urldate=long, date=long, backend=biber]{biblatex}
-\addbibresource{sources.bib}
-
-% General document
-\usepackage[useregional]{datetime2} % Date
-\usepackage{titling} % Allows for multiple titles (info sheet)
-\usepackage{libertinus} % Fonts
-\usepackage[margin=3cm]{geometry} % Custom margins
-\usepackage{parskip} % Removes indentation at start of paragraph, add space between them
-\usepackage[normalem]{ulem} % Underlining
-\useunder{\uline}{\ul}{}
-\usepackage[htt]{hyphenat} % Hyphenation for monospace text
-\usepackage{todonotes} % Add todos
-\usepackage{blindtext} % lorem ipsum, blind text
-
-% Graphics
-\usepackage{graphicx} % Graphics extended
-\usepackage{hyperref} % Refer to things, including URLs
-\usepackage{float} % Allows for H float option ("here")
-\usepackage{caption} % allows for centered figure caption justification
+\usepackage{pagestyle}
+\usepackage{codestyle}
 
@@
 
-\title{\Huge{Machine Learning}\\\LARGE TODO: Title goes here}
+\title{
+    \Huge{Machine Learning} \\ \LARGE 
+    Nature's Quest: Deep Learning Exploration of Global Plant Traits from Images and Geodata
+}
 \author{
@@
 
-\blinddocument
+% \lstinputlisting[firstline=10, lastline=13]{code/sample_script.py}
 
-\include{info_sheet.tex}
+\begin{abstract}
+    % Abstract content goes here
+\end{abstract}
+
+\section{Introduction}
+
+\section{Literature review}
+
+\section{Data processing}
+
+\section{Method}
+
+\subsection{Model design}
+
+\subsection{Model evaluation}
+
+\subsection{Results}
+
+\section{Discussion}
+
+\section{Conclusion}
+
 \end{document}

### 2024-03-10 src/loss.py
stratified k-fold sampling and eval

@@ -1,0 +1,76 @@
+import keras
+import config
+from keras import ops
+
+# Code is in part taken from https://www.kaggle.com/code/awsaf49/planttraits2024-kerascv-starter-notebook 
+# and modified to account for imputing
+
+class R2Loss(keras.losses.Loss):
+    """
+    R2 loss to train the model
+    """
+    def __init__(self, name="r2_loss"):
+        super().__init__(name=name)
+
+    def call(self, y_true, y_pred):
+        """
+        Args:
+            y_true: true values of sample
+            y_pred: predicted values of sample
+        
+        """
+        SS_res = ops.sum(ops.square(y_true - y_pred), axis=0)  
+        SS_tot = ops.sum(
+            ops.square(y_true - ops.mean(y_true, axis=0)), axis=0
+        )  
+        r2_loss = SS_res / (SS_tot + 1e-6)  
+        return ops.mean(r2_loss)  
+    
+class adjusted_R2Loss(keras.losses.Loss):
+    """
+    adjusted_R2Loss as alternative loss
+    """
+    def __init__(self, name="adjusted_r2_loss"):
+        super().__init__(name=name)
+        
+    def call(self, y_true, y_pred):
+        """
+        Args:
+            y_true: true values of sample
+            y_pred: predicted values of sample
+        """
+        SS_res = ops.sum(ops.square(y_true - y_pred), axis=0)  
+        SS_tot = ops.sum(
+            ops.square(y_true - ops.mean(y_true, axis=0)), axis=0
+        )  
+        r2_loss = 1 - SS_res / (SS_tot + 1e-6)
+        df_res = ops.shape(y_true)[0] - config.num_classes - 1
+        df_tot = ops.shape(y_true)[0] - 1
+        adjusted_r2_loss = 1 - (1 - r2_loss) * (df_tot / df_res)
+        return ops.mean(adjusted_r2_loss)
+
+class R2Metric(keras.metrics.Metric):
+    """
+    Use R2 metric to evaluate the model performance after training
+    """
+    def __init__(self, name="r2", **kwargs):
+        super(R2Metric, self).__init__(name=name, **kwargs)
+        self.SS_res = self.add_weight(name="SS_res", shape=(6,), initializer="zeros")
+        self.SS_tot = self.add_weight(name="SS_tot", shape=(6,), initializer="zeros")
[…]

### 2024-03-10 src/databuilder.py
changes

@@ -1,0 +1,130 @@
+import keras_cv
+import keras
+from keras import ops
+import tensorflow as tf
+import cv2
+from config import config
+import os
+
+# code is taken from https://www.kaggle.com/code/awsaf49/planttraits2024-kerascv-starter-notebook 
+
+def build_augmenter():
+    # Define augmentations
+    aug_layers = [
+        keras_cv.layers.RandomBrightness(factor=0.1, value_range=(0, 1)),
+        keras_cv.layers.RandomContrast(factor=0.1, value_range=(0, 1)),
+        keras_cv.layers.RandomSaturation(factor=(0.45, 0.55)),
+        keras_cv.layers.RandomHue(factor=0.1, value_range=(0, 1)),
+        keras_cv.layers.RandomCutout(height_factor=(0.06, 0.15), width_factor=(0.06, 0.15)),
+        keras_cv.layers.RandomFlip(mode="horizontal_and_vertical"),
+        keras_cv.layers.RandomZoom(height_factor=(0.05, 0.15)),
+        keras_cv.layers.RandomRotation(factor=(0.01, 0.05)),
+        keras_cv.layers.RandomGaussianBlur(factor=(0.01, 0.05)),
+    ]
+    
+    # Apply augmentations to random samples
+    aug_layers = [keras_cv.layers.RandomApply(x, rate=0.5) for x in aug_layers]
+    
+    # Build augmentation layer
+    augmenter = keras_cv.layers.Augmenter(aug_layers)
+
+    # Apply augmentations
+    def augment(inp, label):
+        images = inp["images"]
+        aug_data = {"images": images}
+        aug_data = augmenter(aug_data)
+        inp["images"] = aug_data["images"]
+        return inp, label
+    return augment
+
+
+def build_decoder(with_labels=True, target_size=config.image_size):
+    def decode_image(inp):
+        path = inp["images"]
+        
+        # Read jpeg image
+        file_bytes = tf.io.read_file(path)
+        image = tf.io.decode_jpeg(file_bytes)
+        
+        # Resize
+        image = tf.image.resize(image, size=target_size, method="area")
+        
+        # Rescale image
+        image = tf.cast(image, tf.float32)
+        image /= 255.0
+        
+        # Reshape
+        image = tf.reshape(image, [*target_size, 3])
+        
+        inp["images"] = image
+        return inp
+
+    def decode_label(label, num_classes):
+        label = tf.cast(label, tf.float32)
[…]

### 2024-03-25 paper/info_sheet.tex
changes

\title{MLVU information sheet}
\author{}
\date{}
\maketitle

\noindent \textit{Please include this page in your report either at the start or at the end, before the appendix. Do not change the formatting.}

\paragraph{Group number}~

02

\paragraph{Authors}~

\begin{tabular}{r r}
name &student number \\
\hline
Kai Niermann & 1234567 \\
Dávid Miklo & 1234567 \\
Trix Taiclet & 1234567 \\
Conner Dassen & 1234567 \\
Red Kaláb & 1234567 \\
\end{tabular}

\paragraph{Software used} \textit{Describe briefly which software and libraries you relied on. If you built a particular algorithm from scratch, you can explain that here.}

\paragraph{Use of AI tools} \textit{If you used tools like ChatGPT or Github Copilot in any phase of the report, please detail here which tools you used, what parts of the work you used them for and how. Note that if you use an AI tool and don't report it, it's considered fraud, so be detailed. Note also that having an AI tool write for you is not allowed. }

\paragraph{Link to code (optional)} \textit{If you would like to share your code, you can link to your repository here.}

\paragraph{Group disagreements} \textit{If there are any disagreements in the group, you may \textbf{not} remove a student from the author list without their consent. You should bring disagreements to our attention early. As a last resort, you can describe any grievances here}

### 2024-03-25 paper/paper.tex
paper changes

@@
 \section{Introduction}
+% primary paragraph 
+Global warming and more broadly climate change has become a major concern for the world as its effects are becoming more and more apparent \cite{WANG2023100237}.
+Changing weather patterns, especially towards more extreme conditions are causing plants to adapt to new environments \cite{GRAY201664}.
+One method of measuring said adaptation is to look at the traits, that is, properties of a plant that describe how it functions and interacts with the environment.
+These traits include but are not limited to of the plant height, leaf area, but also dry mass, leaf nitrogen content amongst various others.
+Monitoring these traits allows us to gain vital insights into how climate change impacts different ecosystems.
+While somewhat simple manual measurement techniques exist, at scale they are not feasible.
+This is where Convolutional Neural Networks (CNN) come in.
 
+\smallskip
+Through the work demonstrated by Schiller et al \cite{schiller2021deep} we know that CNNs can be used to predict plant traits from images.
+The images used to train this network came from \textit{citizen science photographs} which are images taken by citizens of plants from all across the world using AI plant species identification apps (e.g.
+iNaturalist, Pl@ntNet).
+Citizen science photographs also come with location metadata, which can be used to extract ancillary geodata such as precipitation, temperature, and soil type.
+This geodata can optionally be combined with the images to create a CNN which can potentially learn to extract features from images in conjunction with geodata to predict plant traits.
+
+\smallskip
+For our method we wanted to compare the accuracy of a CNN trained on images alone and using a pretrained backbone (e.g.
+ResNet, EfficientNet, etc.) to a CNN trained on images and geodata.
+We hypothesize that the CNN trained on images and geodata will outperform the CNN trained on images alone.
+Although the geodata could potentially lead the model to learning features not necessarily helpful for accurate prediction of plant traits due to a multitiude of reasons.
+
+% secondary paragraph
 \section{Literature review}
 
-\section{Data processing}
+\section{Method}
 
-\section{Method}
+The key 
+\subsection{Data processing}
 
@@
 
-\subsection{Results}
+\section{Results}
 
@@
 
+\printbibliography
+
 \end{document}

### 2024-03-25 paper/paper.tex
paper updates

@@
 % secondary paragraph
-\section{Literature review}
+% \section{Literature review}
 
@@
 
-The key 
 \subsection{Data processing}
+
+Integral to any machine learning model is the data used to train, validate and ultimate test the model.
+The data used consisted of 3 main components: ancillary geodata from various sources, the main task trait means and auxillary task standard deviations, and the training images of the plants.
+
+\smallskip 
+Upon visual inspection one of the first issues we spotted was a considerable chunk (29.53\%) of missing data missing for the auxillary task standard deviations.
+Through some previous work on the same Kaggle we noticed that the auxillary data was a useful inclusion.
+We can also validate these reports with previous work applying auxillary tasks in CNNs such as the work of Lukas Libel and Marco Kroener which confirmed that the inclusion of an auxillary task with minor relevance to the main task did indeed boost performance \cite{lukaslibel} or Chen et.
+al.
+\cite{pmlr-v80-chen18a} which demonstrated that for very similar or mostly similar auxillary tasks the contribution is generally a positive one.
+Based on this we decided to use a simple k-nearest neighbor (kNN) imputation strategy.
+The rest of the data had all values present.
+
+\smallskip
+The geodata was another point of consideration in the data pre-processing phase.
+Looking at the individual columns we noticed that the instances had features which encoded similar information and thus where likely redundant in the training process.
+Since we are working with a very high dimensional feature space to get some sort of visual verification of this claim we plotted a correlation heatmap for the 3 biggest groups of the geodata, namely the datasets: soil (soil information), MODIS, and VOD.
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=0.9\textwidth]{assets/corr_hm.png}
+    \caption{Correlation heatmap of the 3 biggest groups of geodata}
+\end{figure}
+
+Since these datasets individually can be broken down into groups of features generally talking about a similar topic we can see that the correlation within these groups is quite high.
+Even though previous work \cite{DBLP:journals/corr/abs-2007-00062} has demonstrated the viability of neural network models to learn from high dimensional data, removing redundant features has in numerous instances been shown to improve model performance \cite{chen2022survey}.
+With this in mind we decided to apply a Principal Component Analysis (PCA) to the geodata to reduce the dimensionality of the data.
+We specified that the number of principle components should be such that 95\% of the variance is explained.
+
+\subsection{Data preparation}
 

### 2024-03-26 paper/paper.tex
some more paper structure updates

@@
 \smallskip 
-Upon visual inspection one of the first issues we spotted was a considerable chunk (29.53\%) of missing data missing for the auxillary task standard deviations.
+Upon visual inspection one of the first issues we spotted was a considerable chunk (29.53\%) of data missing for the auxillary task standard deviations.
 Through some previous work on the same Kaggle we noticed that the auxillary data was a useful inclusion.
@@
 \cite{pmlr-v80-chen18a} which demonstrated that for very similar or mostly similar auxillary tasks the contribution is generally a positive one.
-Based on this we decided to use a simple k-nearest neighbor (kNN) imputation strategy.
+Based on this we decided to use a simple k-nearest neighbor (kNN) imputation strategy which has generally proved effective \cite{joel2024performance}.
 The rest of the data had all values present.
@@
 Looking at the individual columns we noticed that the instances had features which encoded similar information and thus where likely redundant in the training process.
-Since we are working with a very high dimensional feature space to get some sort of visual verification of this claim we plotted a correlation heatmap for the 3 biggest groups of the geodata, namely the datasets: soil (soil information), MODIS, and VOD.
+Since we are working with a very high dimensional feature space to get some sort of visual verification of this claim we plotted a correlation heatmap for the 3 biggest groups of the geodata, namely the datasets: SOIL, MODIS, and VOD.
 
@@
 
+% talk about the augmentation 
+Our training and test image instance - consisting of science citizen photographs - where taken by a wide variety of people under a wide variety of conditions.
+This means that the images are of varying quality, resolution, and orientation.
+We can see this just taking a random sample of 5 images.
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=0.9\textwidth]{assets/before_aug_img.png}
+    \caption{Random sample of 5 images before augmentation annotated with their target trait labels}
+\end{figure}
+
+To prevent our model from fitting to noise in the data as a byproduct of how images where taken we decided to apply different keras preprocessing layers to batches concurrently.
+A key benefit of this would be that the model can generalize better to unseen data by learning potentially more meaningful features.
+While this approach of applying augmentations in the preprocessing stage is less flexible than having filters integrated in the model it does reduce model complexity though at the cost of potentially losing some ability to learn more complex features.
+Sampling again we can look at the results of our augmentaions.
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=0.9\textwidth]{assets/after_aug_img.png}
+    \caption{Random sample of 5 images with augmentations: brightness, rotation, contrast, zoom, hue, cutout, flip, blur, saturation}
+\end{figure}
+
+% 5-fold stratified cross-validation + verification and split
+
 \subsection{Model design}
 
+\subsubsection{CNN trained on images}
+
+\subsubsection{CNN trained on images and geodata}
+
 \subsection{Model evaluation}
+
+% evaluation of individual models
+
+% comparative evaluation
 

### 2024-03-26 paper/paper.tex
paper stuff

@@
 % 5-fold stratified cross-validation + verification and split 
+% \cite{Marcot2021} k = 5 good for large sample
+% \cite{kequal10isgood} k ~ 10 generally good choice
+Another method of reducing overfitting we thought relevant to apply here would be k-fold cross validation.
+For the choice of $k$ we chose 5 as for large datasets this has proven effective \cite{Marcot2021} and is generally around the commonly chosen value of 10 \cite{kequal10isgood}.
+Given that the data is from a geographically large enough sample we decided to create the folds in a stratified manner under the assumption the distribution of traits would be reasonably representaitive of the true distribution and thus one the model should learn.
+This is important as it ensures that the model is trained on a representative sample of the data in each fold and in turn has its predictions distributed in a similar manner to the assumed true distribution of the traits.
+So we reduce the risk of overfitting to the training data while also ensuring the model is trained using a representaitive sample of targets in each fold.
+To validate that the distributions of the target traits are indeed similar we averaged the distributions of the target traits in each fold and overlayed this on the original distribution of trait means and normalized them to the same scale.
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=1\textwidth]{assets/distribution_match_folds.png}
+    \caption{Comparison of the average trait distributions in each fold with the original distribution of trait means ; (blue) normalized distribution of the trait means across the entire dataset ; (yellow) average normalized distribution of the trait means in each fold}
+\end{figure}
 
@@
 \subsubsection{CNN trained on images}
+
+% give an overview of the model 
+    % activation function 
+    % batch size
+    % optimizer 
+    % epochs
+Our image CNN is a multi-output model which takes in a tensor consisting of red, green, and blue channels with an image size of 224x224 pixels.
+The model consists of a pretrained backbone; in our instance we compared \textit{efficientnetv2\_b2\_imagenet}, \textit{efficientnetv2\_s\_imagenet}, and \textit{resnet50\_imagenet}.
+The backbone outputs then feed into a global average pooling layer, followed by 2 dense layers and a dropout layer connected to the two dense output layers for the main and auxillary task.
+The hidden layers use a ReLU activation function, the outputs use no activation function as the targets are continuous.
+The model is trained using the Adam optimizer with a learning rate of 1e-4, a batch size of 48 and for 12 epochs.
+Additionally we used the $R^2$ metric for both loss and evaluation of the model performance as this is a regression problem.
+Finally we used step-based for the learning rate schedule which is defined as follows
+\[
+    \text{lr} = \text{lr}_\text{max} \times d^{\floor*{\frac{e_d}{r}}}
+\]
+Where $e_d$ is the current epoch, $r$ is the step size (=2), and $d$ is the decay rate.
+
+% talk about the backbones 
+    % ResNet 
+    % EfficientNet b2 
+    % EfficientNet s
+\smallskip
+Residual Network (ResNet) \cite{he2016identity} introduced in 2016 is a deep learning model that is able to learn features from images.
+Most famously its known for mitigating the vanishing gradient problem through the introduction of identiy skip connections.
+We are using \text{ResNet50} which was trained on imagenet and consists of 48 convolution layers, in addition to a MaxPool and AveragePool layer.
+EfficientNetV2 \cite{tan2021efficientnetv2} is a family of CNNs from 2021 that have faster speeds than many previous models primarily by increasing image size during training and incorporating adaptive regularization to compensate for any potential accuracy loss.
+We are using \textit{efficientnetv2\_b2\_imagenet} and \textit{efficientnetv2\_s\_imagenet} which are pretrained on imagenet, with the ladder being a smaller (fewer parameters) version of the former.
+Comparing these backbones in a table we can get some overview of how they stack up as seen in \ref{tab:backbone_comparison}
+
+To compare these backbones we trained 3 different models with the same architecture but different backbones.
+We also specified a contribution of 0.3 to the final loss of the auxillary task (as opposed to 1.0 for the main task) as to limit its impacts on the direction of training.
+A visual overview of the different models can be seen in \ref{fig:models_overview}.
+
+\begin{table}[!h]
+    % ref for this table
+
+    \centering
+    \begin{tabular}{@{}llll@{}}
+    \toprule
+    & EfficientNetV2 b2 & EfficientNetV2 s & ResNet50 \\ \midrule
+    parameters              & 8.77M             & 20.33M           & 23.56M   \\
+    imagenet top 5 accuracy & 94.9\%            & 96.7\%           & (missing)       \\ \bottomrule
+\end{tabular}
+\caption{Comparison of the different backbones and published top 5 accuracy on imagenet for the specific models}
+\label{tab:backbone_comparison}
+\end{table}
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=1\textwidth]{assets/different_models.png}
+    \caption{Visual overview of the 3 models with different backbones}
+    \label{fig:models_overview}
+\end{figure}
 

### 2024-03-27 src/loss.py
updates

@@ -9,7 +9,7 @@ class R2Loss(keras.losses.Loss):
     """
     R2 loss to train the model
     """
-    def __init__(self, name="r2_loss"):
+    def __init__(self, name="r2_loss", reduction="sum_over_batch_size"):
         super().__init__(name=name)
 
     def call(self, y_true, y_pred):

### 2024-03-27 paper/paper.tex

@@
 Additionally we used the $R^2$ metric for both loss and evaluation of the model performance as this is a regression problem.
-Finally we used step-based for the learning rate schedule which is defined as follows
+Finally we used step-based for the learning rate schedule which was defined as follows
 \[
@@
 
+\subsubsection{Image CNN tuning}
+
+% model x had the best loss in the end 
+Looking at the loss metrics for the main \textit{head} (main) task we can observe that the network using the <backbone> backbone proved most accurate.
+Having an $R^2$ of <score> at its best epoch measured on the training data.
+Though further tuning could possibly improve our performance.
+Firstly we tested out different learning rate schedulers, specifically \textit{cosine annealing} \cite{cosineannealing} and \textit{exponential decay} using the model with the best backbone.
+The results of this can be seen in \ref{fig:lr_schedulers}.
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=0.3\textwidth]{assets/sample_image.jpeg}
+    \caption{Comparison of the different learning rate schedulers}
+    \label{fig:lr_schedulers}
+\end{figure}
+
+Using these results we once again take the model with the best learning rate scheduler, finally we thought it would make sense to see if the trainability of the backbone had any effects on the models prediction capabilities.
+While fine-tuning does allow our model to adapt the backbone for our specific task it does also risk overfitting to our training data though the large amount of data we have should limit this risk somewhat \cite{Ying_2019}.
+So in \ref{fig:trainable_backbones} we compared the result of our model and without training enabled for the <backbone> backbone.
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=0.3\textwidth]{assets/sample_image.jpeg}
+    \caption{Comparison of the model error over epochs for a trainable and non-trainable backbone}
+\end{figure}
+
+% we wanted to test out : different lr, backbone training on vs off 
+    % test out cosine annealing 
+    % test out exp 
+    % plot step vs exp vs cosine annealing
+
+    % test out backbone training off for best lr schedule 
+    % test out backbone training on for best lr schedule
+
 \subsubsection{CNN trained on images and geodata}
+
+% give an overview of the model
+    % mention mainly the geodata model 
+    % show a plot of the model
 
@@
 
+% show intermediate activations and maybe comment on those
+
 % comparative evaluation
@@
 
+% give a simple summary of the results
+
 \section{Discussion}
 
+% discuss the results
+    % why did we see what we saw 
+    % what does this mean for the main task 
+    % what are some of the limitations of our approach
+
 \section{Conclusion}
+
+% give some future direction for things to go into
 

### 2024-03-28 paper/paper.tex
new additions

--- 2024-03-26 paper/paper.tex
@@
 % primary paragraph 
-Global warming and more broadly climate change has become a major concern for the world as its effects are becoming more and more apparent \cite{WANG2023100237}.
+Global warming and more broadly climate change have become a major concern for the world as its effects are becoming more and more apparent \cite{WANG2023100237}.
 Changing weather patterns, especially towards more extreme conditions are causing plants to adapt to new environments \cite{GRAY201664}.
 One method of measuring said adaptation is to look at the traits, that is, properties of a plant that describe how it functions and interacts with the environment.
-These traits include but are not limited to of the plant height, leaf area, but also dry mass, leaf nitrogen content amongst various others.
+These traits include but are not limited to the plant height, leaf area, but also dry mass, and leaf nitrogen content amongst various others.
 Monitoring these traits allows us to gain vital insights into how climate change impacts different ecosystems.
@@
 \smallskip
-For our method we wanted to compare the accuracy of a CNN trained on images alone and using a pretrained backbone (e.g.
+For our method, we wanted to compare the accuracy of a CNN trained on images alone and using a pre-trained backbone (e.g.
 ResNet, EfficientNet, etc.) to a CNN trained on images and geodata.
 We hypothesize that the CNN trained on images and geodata will outperform the CNN trained on images alone.
-Although the geodata could potentially lead the model to learning features not necessarily helpful for accurate prediction of plant traits due to a multitiude of reasons.
+Although the geodata could potentially lead the model to learning features not necessarily helpful for accurate prediction of plant traits due to a multitude of reasons.
 
@@
 
-Integral to any machine learning model is the data used to train, validate and ultimate test the model.
-The data used consisted of 3 main components: ancillary geodata from various sources, the main task trait means and auxillary task standard deviations, and the training images of the plants.
+Integral to any machine learning model is the data used to train, validate, and ultimately test the model.
+The data used consisted of 3 main components: ancillary geodata from various sources, the main task trait means and auxiliary task standard deviations, and the training images of the plants.
 
 \smallskip 
-Upon visual inspection one of the first issues we spotted was a considerable chunk (29.53\%) of data missing for the auxillary task standard deviations.
-Through some previous work on the same Kaggle we noticed that the auxillary data was a useful inclusion.
-We can also validate these reports with previous work applying auxillary tasks in CNNs such as the work of Lukas Libel and Marco Kroener which confirmed that the inclusion of an auxillary task with minor relevance to the main task did indeed boost performance \cite{lukaslibel} or Chen et.
+Upon visual inspection one of the first issues we spotted was a considerable chunk (29.53\%) of data missing for the auxiliary task standard deviations.
+Through some previous work on the same Kaggle, we noticed that the auxiliary data was a useful inclusion.
+We can also validate these reports with previous work applying auxiliary tasks in CNNs such as the work of Lukas Libel and Marco Kroener which confirmed that the inclusion of an auxiliary task with minor relevance to the main task did indeed boost performance \cite{lukaslibel} or Chen et.
 al.
-\cite{pmlr-v80-chen18a} which demonstrated that for very similar or mostly similar auxillary tasks the contribution is generally a positive one.
+\cite{pmlr-v80-chen18a} which demonstrated that for very similar or mostly similar auxiliary tasks the contribution is generally a positive one.
 Based on this we decided to use a simple k-nearest neighbor (kNN) imputation strategy which has generally proved effective \cite{joel2024performance}.
@@
 The geodata was another point of consideration in the data pre-processing phase.
-Looking at the individual columns we noticed that the instances had features which encoded similar information and thus where likely redundant in the training process.
+Looking at the individual columns we noticed that the instances had features that encoded similar information and thus were likely redundant in the training process.
 Since we are working with a very high dimensional feature space to get some sort of visual verification of this claim we plotted a correlation heatmap for the 3 biggest groups of the geodata, namely the datasets: SOIL, MODIS, and VOD.
@@
 Even though previous work \cite{DBLP:journals/corr/abs-2007-00062} has demonstrated the viability of neural network models to learn from high dimensional data, removing redundant features has in numerous instances been shown to improve model performance \cite{chen2022survey}.
-With this in mind we decided to apply a Principal Component Analysis (PCA) to the geodata to reduce the dimensionality of the data.
-We specified that the number of principle components should be such that 95\% of the variance is explained.
+With this in mind, we decided to apply a Principal Component Analysis (PCA) to the geodata to reduce the dimensionality of the data.
+We specified that the number of principal components should be such that 95\% of the variance is explained.
 
@@
 % talk about the augmentation 
-Our training and test image instance - consisting of science citizen photographs - where taken by a wide variety of people under a wide variety of conditions.
+Our training and test image instance - consisting of science citizen photographs - were taken by a wide variety of people under a wide variety of conditions.
 This means that the images are of varying quality, resolution, and orientation.
-We can see this just taking a random sample of 5 images.
+We can see this just by taking a random sample of 5 images.
 \begin{figure}[!h]
@@
 
-To prevent our model from fitting to noise in the data as a byproduct of how images where taken we decided to apply different keras preprocessing layers to batches concurrently.
+To prevent our model from fitting to noise in the data as a byproduct of how images were taken we decided to apply different keras preprocessing layers to batches concurrently.
 A key benefit of this would be that the model can generalize better to unseen data by learning potentially more meaningful features.
-While this approach of applying augmentations in the preprocessing stage is less flexible than having filters integrated in the model it does reduce model complexity though at the cost of potentially losing some ability to learn more complex features.
-Sampling again we can look at the results of our augmentaions.
+While this approach of applying augmentations in the preprocessing stage is less flexible than having filters integrated into the model it does reduce model complexity though at the cost of potentially losing some ability to learn more complex features by being able to dynamically adjust the filters.
+Sampling again we can look at the results of our augmentations.
 
@@
 % \cite{kequal10isgood} k ~ 10 generally good choice
-Another method of reducing overfitting we thought relevant to apply here would be k-fold cross validation.
+
+\smallskip 
+In addition to augmentation we noted that - though not too common - there was still a presence of certain substantial outliers in our data.
+Through visual inspection a common reason seems to have been specifically how close an indivdiual was when they took an image.
+As the training data itself was labelled by a machine learning model in a supervised fashion its evident perspective was not something the intial model had learned.
+As this scaling issue is a byproduct of the inital model which labelled the data and not any true property of the data we decided to remove these outliers.
+We did this by constraining the target distributions to the interval $[0.0001, 0.9999]$ and then removing any instances that were outside of this interval.
+
+\smallskip
+Another method of reducing overfitting we thought relevant to apply here would be k-fold cross-validation.
 For the choice of $k$ we chose 5 as for large datasets this has proven effective \cite{Marcot2021} and is generally around the commonly chosen value of 10 \cite{kequal10isgood}.
-Given that the data is from a geographically large enough sample we decided to create the folds in a stratified manner under the assumption the distribution of traits would be reasonably representaitive of the true distribution and thus one the model should learn.
-This is important as it ensures that the model is trained on a representative sample of the data in each fold and in turn has its predictions distributed in a similar manner to the assumed true distribution of the traits.
-So we reduce the risk of overfitting to the training data while also ensuring the model is trained using a representaitive sample of targets in each fold.
-To validate that the distributions of the target traits are indeed similar we averaged the distributions of the target traits in each fold and overlayed this on the original distribution of trait means and normalized them to the same scale.
+Given that the data is from a geographically large enough sample we decided to create the folds in a stratified manner under the assumption the distribution of traits would be reasonably representative of the true distribution and thus one the model should learn.
+This is important as it ensures that the model is trained on a representative sample of the data in each fold and in turn, has its predictions distributed in a similar manner to the assumed true distribution of the traits.
+So we reduce the risk of overfitting to the training data while also ensuring the model is trained using a representative sample of targets in each fold.
+To validate that the distributions of the target traits are indeed similar we averaged the distributions of the target traits in each fold overlayed this on the original distribution of trait means and normalized them to the same scale.
 
@@
     % epochs
-Our image CNN is a multi-output model which takes in a tensor consisting of red, green, and blue channels with an image size of 224x224 pixels.
-The model consists of a pretrained backbone; in our instance we compared \textit{efficientnetv2\_b2\_imagenet}, \textit{efficientnetv2\_s\_imagenet}, and \textit{resnet50\_imagenet}.
-The backbone outputs then feed into a global average pooling layer, followed by 2 dense layers and a dropout layer connected to the two dense output layers for the main and auxillary task.
+Our image CNN is a multi-output model that takes in a tensor consisting of red, green, and blue channels with an image size of 224x224 pixels.
+The model consists of a trained backbone; in our instance, we compared \textit{efficientnetv2\_b2\_imagenet}, \textit{efficientnetv2\_s\_imagenet}, and \textit{resnet50\_imagenet}.
+The backbone outputs then feed into a global average pooling layer, followed by 2 dense layers and a dropout layer connected to the two dense output layers for the main and auxiliary tasks.
 The hidden layers use a ReLU activation function, the outputs use no activation function as the targets are continuous.
-The model is trained using the Adam optimizer with a learning rate of 1e-4, a batch size of 48 and for 12 epochs.
-Additionally we used the $R^2$ metric for both loss and evaluation of the model performance as this is a regression problem.
-Finally we used step-based for the learning rate schedule which is defined as follows
+The model is trained using the Adam optimizer with a learning rate of 1e-4, a batch size of 48, and for 12 epochs.
+Additionally, we used the $R^2$ metric for both loss and evaluation of the model performance as this is a regression problem.
+Finally, we used step-based for the learning rate schedule which was defined as follows
 \[
@@
 Residual Network (ResNet) \cite{he2016identity} introduced in 2016 is a deep learning model that is able to learn features from images.
-Most famously its known for mitigating the vanishing gradient problem through the introduction of identiy skip connections.
+Most famously it's known for mitigating the vanishing gradient problem through the introduction of identity skip connections.
 We are using \text{ResNet50} which was trained on imagenet and consists of 48 convolution layers, in addition to a MaxPool and AveragePool layer.
 EfficientNetV2 \cite{tan2021efficientnetv2} is a family of CNNs from 2021 that have faster speeds than many previous models primarily by increasing image size during training and incorporating adaptive regularization to compensate for any potential accuracy loss.
-We are using \textit{efficientnetv2\_b2\_imagenet} and \textit{efficientnetv2\_s\_imagenet} which are pretrained on imagenet, with the ladder being a smaller (fewer parameters) version of the former.
+We are using \textit{efficientnetv2\_b2\_imagenet} and \textit{efficientnetv2\_s\_imagenet} which are trained on imagenet, with the ladder being a smaller (fewer parameters) version of the former.
 Comparing these backbones in a table we can get some overview of how they stack up as seen in \ref{tab:backbone_comparison}
@@
 To compare these backbones we trained 3 different models with the same architecture but different backbones.
-We also specified a contribution of 0.3 to the final loss of the auxillary task (as opposed to 1.0 for the main task) as to limit its impacts on the direction of training.
+We also specified a contribution of 0.3 to the final loss of the auxiliary task (as opposed to 1.0 for the main task) as to limit its impacts on the direction of training.
 A visual overview of the different models can be seen in \ref{fig:models_overview}.
@@
 
+\subsubsection{Image CNN tuning}
+
+% model x had the best loss in the end 
+Looking at the results of the inital choice of hyperparameters for the different models the first thing we noticed was the similar behavior of all the $R^2$ values.
+The coefficient of determination is defined as follows
+
+\[
+    R^2 = 1 - \frac{\sum_{i=1}^{n} (y_i - \hat{y}_i)^2}{\sum_{i=1}^{n} (y_i - \bar{y})^2} = 1 - \frac{SS_{res}}{SS_{tot}}  
+\]
+Where $SS_{res}$ is the residual sum of squares and $SS_{tot}$ is the total sum of squares.
+One observation in our results was that for all instances we had a negative $R^2$, implying that $SS_{res} > SS_{tot}$ meaning our model preforms worse than the baseline at $SS_{res} = SS_{tot}$.
+This is a clear indication that the model is evidently learning nothing from the data, numerious reasons could be the cause of this.
+
+\smallskip 
+One of the first things we decided to test was the learning rate schedule.
+We tested out 3 different schedules: step-based (baseline), exponential, and cosine annealing.
+Due to the results of the intial training not being conclusive in regards to which model is better we decided to to simply continue with the model using the \textit{efficientnetv2\_b2\_imagenet} backbone as it a high top 5 accuracy while also having fewer parameters, reason being that the source of the problem is likely not the backbone rather but rather to do with the model architecture surrounding this and the hyperparameter choices.
+
+% we wanted to test out : different lr, backbone training on vs off 
+    % test out cosine annealing 
+    % test out exp 
+    % plot step vs exp vs cosine annealing
+
+    % test out backbone training off for best lr schedule 
+    % test out backbone training on for best lr schedule
+
 \subsubsection{CNN trained on images and geodata}
+
+% give an overview of the model
+    % mention mainly the geodata model 
+    % show a plot of the model
 

### 2024-03-28 src/combined_model.py

@@ -1,0 +1,41 @@
+import keras_cv
+import tensorflow.keras as keras
+from tensorflow.keras import layers
+from loss import R2Loss, R2Metric
+from keras.metrics import Accuracy
+import config
+
+def cModel(config, feature_cols, img_model, model_name=None):
+    
+    feature_in = keras.Input(shape=(len(feature_cols),), name="features")
+    
+    # tabular input branch
+    dense = layers.Dense(326, activation="selu")(feature_in)
+    dense = layers.Dense(64, activation="selu")(dense)
+    dropout = layers.Dropout(0.1)(dense)
+    
+    concat = layers.Concatenate()([img_model[0], dropout])
+    
+    # output layers
+    head_out = layers.Dense(config.num_classes, activation=None, name="head")(concat)
+    aux_head_out = layers.Dense(config.aux_num_classes, activation="relu", name="aux_head")(concat)
+
+    model = keras.models.Model(inputs=[img_model[1], feature_in], outputs={"head": head_out, "aux_head": aux_head_out})
+    
+    if model_name:
+        model.name = model_name
+
+    # Compile model
+    print("Compiling model")
+    model.compile(
+        optimizer=keras.optimizers.Adam(learning_rate=1e-4),
+        loss={"head": R2Loss(), "aux_head": R2Loss()},
+        loss_weights={"head": 1.0, "aux_head": 0.3},
+        metrics={"head": R2Metric()},
+    )
+
+    return model
+
+def model_summary(model):
+    model.build((None, *config.image_size, 3))
+    model.summary()

### 2024-03-28 src/config.py

@@ -10,7 +10,6 @@ class config:
     epochs = 12                 # Training epochs
     batch_size = 48             # Batch size
     lr_mode = "step"            # LR scheduler mode from one of "cos", "step", "exp"
-    drop_remainder = True       # Drop incomplete batches
     num_classes = 6             # Number of classes in the dataset
     num_folds = 5               # Number of folds to split the dataset
     fold = 0                    # Which fold to set as validation data

### 2024-03-28 src/databuilder.py

@@ -19,7 +19,6 @@ def build_augmenter():
         keras_cv.layers.RandomFlip(mode="horizontal_and_vertical"),
         keras_cv.layers.RandomZoom(height_factor=(0.05, 0.15)),
         keras_cv.layers.RandomRotation(factor=(0.01, 0.05)),
-        keras_cv.layers.RandomGaussianBlur(3, factor=(0.01, 0.05)),
     ]
     
     # Apply augmentations to random samples

### 2024-03-28 src/model.py

@@ -5,7 +5,7 @@ from loss import R2Loss, R2Metric
 from keras.metrics import Accuracy
 import config
 
-def Model(config, slct_backbone):
+def Model(config, slct_backbone, train_backbone=True, model_name=None, partial=False):
     backbones = {
         "efficientnetv2_b2_imagenet": keras_cv.models.EfficientNetV2B2Backbone,
         "efficientnetv2_s_imagenet": keras_cv.models.EfficientNetV2SBackbone,
@@ -17,24 +17,30 @@ def Model(config, slct_backbone):
 
     # Branch for image input
     backbone = backbones[slct_backbone](slct_backbone)
+    backbone.trainable = train_backbone
     backbone_output = backbone(img_input)
-    gavrg1 = layers.GlobalAveragePooling2D()(backbone_output)
-    dense1 = layers.Dense(1024, activation="relu")(gavrg1)
-    dense2 = layers.Dense(512, activation="relu")(dense1)
-    dropout = layers.Dropout(0.2)(dense2)
+    avrgpool        = layers.GlobalAveragePooling2D()(backbone_output)
+    dense           = layers.Dense(1024, activation="relu")(avrgpool)
+    dense           = layers.Dense(512, activation="relu")(dense)
+    dropout         = layers.Dropout(0.2)(dense)
+
+    if partial:
+        return dropout, img_input
 
     # Output layers
     head_out = layers.Dense(config.num_classes, activation=None, name="head")(dropout)
     aux_head_out = layers.Dense(config.aux_num_classes, activation="relu", name="aux_head")(dropout)
 
     model = keras.Model(inputs=img_input, outputs={"head": head_out, "aux_head": aux_head_out})
+    if model_name:
+        model.name = model_name
 
     # Compile model
     print("Compiling model")
     model.compile(
         optimizer=keras.optimizers.Adam(learning_rate=1e-4),
         loss={"head": R2Loss(), "aux_head": R2Loss()},
-        loss_weights={"head": 1.0, "aux_head": 0.3},
+        loss_weights={"head": 1.0, "aux_head": 0.15},
         metrics={"head": R2Metric()},
     )
 

### 2024-03-29 paper/paper.tex
bunch of changes

@@
 Through some previous work on the same Kaggle, we noticed that the auxiliary data was a useful inclusion.
-We can also validate these reports with previous work applying auxiliary tasks in CNNs such as the work of Lukas Libel and Marco Kroener which confirmed that the inclusion of an auxiliary task with minor relevance to the main task did indeed boost performance \cite{lukaslibel} or Chen et.
+We can also validate these reports with previous work applying auxiliary tasks in CNNs such as the work of Lukas Libel and Marco Kroener which found that the inclusion of an auxiliary task with minor relevance to the main task did indeed boost performance \cite{lukaslibel} and Chen et.
 al.
@@
 
-Since these datasets individually can be broken down into groups of features generally talking about a similar topic we can see that the correlation within these groups is quite high.
+Since these datasets individually can be broken down into groups of features generally talking about a similar topic (e.g.
+precipitation metrics, temperature metrics, etc) we can see that the correlation within these groups is quite high.
 Even though previous work \cite{DBLP:journals/corr/abs-2007-00062} has demonstrated the viability of neural network models to learn from high dimensional data, removing redundant features has in numerous instances been shown to improve model performance \cite{chen2022survey}.
@@
 \smallskip
-Another method of reducing overfitting we thought relevant to apply here would be k-fold cross-validation.
-For the choice of $k$ we chose 5 as for large datasets this has proven effective \cite{Marcot2021} and is generally around the commonly chosen value of 10 \cite{kequal10isgood}.
+Another method of reducing overfitting we thought relevant to apply here would be k-fold cross-validation for the data split.
+We chose $k=5$ as for large datasets this has proven effective \cite{Marcot2021} and is generally around the commonly chosen value of 10 \cite{kequal10isgood}, where the 0th fold was designated as the validation set.
 Given that the data is from a geographically large enough sample we decided to create the folds in a stratified manner under the assumption the distribution of traits would be reasonably representative of the true distribution and thus one the model should learn.
@@
 % model x had the best loss in the end 
-Looking at the results of the inital choice of hyperparameters for the different models the first thing we noticed was the similar behavior of all the $R^2$ values.
+Looking at the results in \ref{fig:different_backbones} of the inital choice of hyperparameters for the different models the first thing we noticed was similar behavior of all the $R^2$ values in terms of their sign.
 The coefficient of determination is defined as follows
@@
 One observation in our results was that for all instances we had a negative $R^2$, implying that $SS_{res} > SS_{tot}$ meaning our model preforms worse than the baseline at $SS_{res} = SS_{tot}$.
-This is a clear indication that the model is evidently learning nothing from the data, numerious reasons could be the cause of this.
+This is a clear indication that the model is fitting against the data, numerious reasons could be the cause of this.
+Due to the results of the intial training not being conclusive in regards to which model is better we decided to to simply continue with the model using the \textit{efficientnetv2\_b2\_imagenet} backbone as it has a high top 5 accuracy while also having fewer parameters.
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=1\textwidth]{assets/different_backbones.png}
+    \caption{Validation loss for primary task with different backbones}
+    \label{fig:different_backbones}
+\end{figure}
+
+While both of the EfficientNet backbones exhibit a similar behavior as they come from the same family of models; this being a fast convergence though also an early plateu at a more suboptimal loss.
+The ResNet50 backbone on the other hand has a slower convergence, but a more consisestent decrease in loss (increase in $R^2$).
+This could indicate that the ResNet50 backbone has a greater potential to learn from the data albeit in a slower manner.
 
@@
 We tested out 3 different schedules: step-based (baseline), exponential, and cosine annealing.
-Due to the results of the intial training not being conclusive in regards to which model is better we decided to to simply continue with the model using the \textit{efficientnetv2\_b2\_imagenet} backbone as it a high top 5 accuracy while also having fewer parameters, reason being that the source of the problem is likely not the backbone rather but rather to do with the model architecture surrounding this and the hyperparameter choices.
+The results demonstrated in \ref{fig:lr_schedules} show that the cosine annealing and exponential schedule had a slower convergence than the step-based schedule.
+This is likely due to the fact that the step-based schedule is more aggressive in its learning rate decay.
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=1\textwidth]{assets/lr_schedules.png}
+    \caption{Comparison of different learning rate schedules}
+    \label{fig:lr_schedules}
+\end{figure}
 
@@
 
+Another point of consideration we had was the training of the backbone.
+We wanted to test out whether there was any difference in $R^2$ whether we allowed the model to adapt the weights of the backbone or not.
+The results of this can be seen in \ref{fig:backbone_training} in both of these instances we see the model converging at around the 2nd epoch and hitting a plateu of $\approx -0.03 R^2$.
+This is likely due to the fact that that the backbone requires little additional training to learn the features from the data and generalizes well to this task.
+
     % test out backbone training off for best lr schedule 
@@
 
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=0.7\textwidth]{assets/train_vs_notrain.png}
+    \caption{Comparison of of models with backbone training on and off}
+    \label{fig:backbone_training}
+\end{figure}
+
 \subsubsection{CNN trained on images and geodata}
+
+To construct the CNN trained on images and geodata we first created the two branches of the model, one to take in the image data and reduce it to a 1-tensor and similarly one to extract features from the tabular data.
+The two branches where then concatenated and fed to the main and auixallary task dense output layers.
+As the only difference here is the additional branch for the geodata we kept the hyperparameters the same to get a fair comparison between both approaches.
+A visual overview of the model can be seen in \ref{fig:geodata_model}.
+
+\begin{figure}[!h]
+    \centering 
+    \includegraphics[width=0.8\textwidth]{assets/img_geo_model.png}
+    \caption{Model architecture of the CNN trained on images and geodata}
+    \label{fig:geodata_model}
+\end{figure}
 
@@
 
-\subsection{Model evaluation}
+\subsection{Model comparative evaluation}
 
@@
 
-% show intermediate activations and maybe comment on those
+For the final comparison we first looked at the basic evaluation metrics of either model.
+The results of this can be seen in \ref{fig:basic_evaluation}.
+We can see that the model trained on images and geodata has a higher $R^2$ suggesting a relative improvement in the prediction capability over just the image model which has a best recorded $R^2$ of <number>.
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=0.7\textwidth]{assets/fin_img_vs_geo.png}
+    \caption{Basic evaluation metrics of the two models}
+    \label{fig:basic_evaluation}
+\end{figure}
+
+\smallskip
 
@@
 % give a simple summary of the results
+One potential reason why the combined model performs better is simply that certain patterns are easier to learn from the geodata than from the images.
+Take the trait \textit{Stem specific density} for example, this is a trait that is likely to be more defined as a function of the environment \cite{plants8030065} rather than any visual cues.
+The same applies to the trait \textit{Leaf nitrogen content} which has been shown to be associated with abiotic stress (e.g.
+heat, cold, drought, etc.) \cite{Ye2022} and thus is likely to be more easily learned from the geodata.
 
@@
 % discuss the results
-    % why did we see what we saw 
-    % what does this mean for the main task 
-    % what are some of the limitations of our approach
+The prediction of plant traits from images and geodata is a complex task that requires a CNN to learn and extract featues from both images and geodata.
+The results of our work show that a CNN trained on images and geodata outperforms a CNN trained on images alone.
+Though neither of the models elaborated here is necessarily a good predictior, one of the main reasons for this is likely the complexity and nuance of the task.
+Though in addition there are various ways our method could be improved upon.
+
+Hyperparameter tuning is one of the most important aspects of training a machine learning model.
+In our work we only tested a few different learning rate schedules and backbones.
+Random search - which as been demonstrated to be more effective than grid search \cite{JMLR:v13:bergstra12a} - would likely have been a better choice, albeit more computationally expensive.
+In addition we predominantly looked at hyperparameters, but varying the model architecture could have also been a benifical direction.
+A larger model could potentially learn more nuanced features in the data.
+
+The training data itself can also be possibly modified to improve the model.
+Certain plant traits rely on some sense of relative scale to be elicited accurately (e.g.
+we know that a tree is bigger than flower regardless of image angle), this scale is easily lost due to the variability in how people take photos.
+One way of mitigating this could be to incorporate scale normalization through Depth Estimation, previous work \cite{Ummenhofer_2017} demonstrated the ability to train CNNs on the DIODE (Dense Indoor/Outdoor DEpth) \cite{vasiljevic2019diode} dataset to estimate depth.
+The learned depth information of the images could be used then as a moderator to scale the images to a common scale possibly improving the prediction capabilities.
+
+How adaptable pretrained models are to new tasks is also is also something that could be further explored.
+Kunze et al.
+\cite{kunze2017transfer} answered this question by looking at the weight difference distribution per layer early and late into training.
+What they found was that for models trained on a similar task to the target domain only had a minor change in the weight distribution.
+This reasoning could possibly be applied to more accurately find the optimal backbone.
+Additionally the use of a pretrained model could be further explored by looking at the effect of the number of layers frozen during training.
 

### 2024-03-29 src/config.py

@@ -7,7 +7,7 @@ class config:
     verbose = 1                 # Verbosity
     seed = 42                   # Random seed
     image_size = [224, 224]     # Input image size
-    epochs = 12                 # Training epochs
+    epochs = 20                 # Training epochs
     batch_size = 48             # Batch size
     lr_mode = "step"            # LR scheduler mode from one of "cos", "step", "exp"
     num_classes = 6             # Number of classes in the dataset

### 2024-03-29 src/model.py

@@ -40,7 +40,7 @@ def Model(config, slct_backbone, train_backbone=True, model_name=None, partial=F
     model.compile(
         optimizer=keras.optimizers.Adam(learning_rate=1e-4),
         loss={"head": R2Loss(), "aux_head": R2Loss()},
-        loss_weights={"head": 1.0, "aux_head": 0.15},
+        loss_weights={"head": 1.0, "aux_head": 0.3},
         metrics={"head": R2Metric()},
     )
 

### 2024-03-29 paper/paper.tex
result plots

@@
 We hypothesize that the CNN trained on images and geodata will outperform the CNN trained on images alone.
-Although the geodata could potentially lead the model to learning features not necessarily helpful for accurate prediction of plant traits due to a multitude of reasons.
+With our results we hope to provide a better understanding of the key factors that are needed to predict plant traits.
 
@@
 Integral to any machine learning model is the data used to train, validate, and ultimately test the model.
-The data used consisted of 3 main components: ancillary geodata from various sources, the main task trait means and auxiliary task standard deviations, and the training images of the plants.
+The data used consisted of 3 main components: ancillary geodata from various sources, the main task trait means and auxiliary task standard deviations, and the training ima ges of the plants.
 
@@
 We tested out 3 different schedules: step-based (baseline), exponential, and cosine annealing.
-The results demonstrated in \ref{fig:lr_schedules} show that the cosine annealing and exponential schedule had a slower convergence than the step-based schedule.
-This is likely due to the fact that the step-based schedule is more aggressive in its learning rate decay.
+The results demonstrated in \ref{fig:lr_schedules} show that the cosine annealing and exponential schedule had a slower convergence than the step-based schedule and a more erratic behavior.
 
@@
     \centering
-    \includegraphics[width=1\textwidth]{assets/lr_schedules.png}
+    \includegraphics[width=0.4\textwidth]{assets/lr_schedule_diffs.png}
     \caption{Comparison of different learning rate schedules}
@@
     \includegraphics[width=0.7\textwidth]{assets/train_vs_notrain.png}
-    \caption{Comparison of of models with backbone training on and off}
+    \caption{Comparison of models with backbone training on (right) and off (left)}
     \label{fig:backbone_training}
@@
     \includegraphics[width=0.8\textwidth]{assets/img_geo_model.png}
-    \caption{Model architecture of the CNN trained on images and geodata}
+    \caption{(right) Model architecture of the CNN trained on images and geodata; (left) Model architecture of the CNN trained on images alone}
     \label{fig:geodata_model}
@@
 The prediction of plant traits from images and geodata is a complex task that requires a CNN to learn and extract featues from both images and geodata.
-The results of our work show that a CNN trained on images and geodata outperforms a CNN trained on images alone.
+The results of our work show that a CNN trained on just image outpreforms one trained on images and geodata.
 Though neither of the models elaborated here is necessarily a good predictior, one of the main reasons for this is likely the complexity and nuance of the task.
-Though in addition there are various ways our method could be improved upon.
+We now propose various directions for future work to improve the prediction capabilities of the models.
 
@@
 \cite{kunze2017transfer} answered this question by looking at the weight difference distribution per layer early and late into training.
-What they found was that for models trained on a similar task to the target domain only had a minor change in the weight distribution.
+What they found was that models trained on a similar task to the target domain only had a minor change in the weight distribution.
 This reasoning could possibly be applied to more accurately find the optimal backbone.
@@
 \section{Conclusion}
-
-% give some future direction for things to go into
+To conclude we noticed that first and foremost the task to predict pant traits is a hard one.
+It requires a careful consideration during every step of the process from understanding the data to how a neural network architecture will learn from it.
+From our results we observed that a model trained on images and ancillary geodata converges faster but does not necessarily have a higher $R^2$ than a model trained on images alone.
+In the final tests run these results were also observed with our image model having a higher $R^2$ than the combined model.
+Though both models still have a negative $R^2$ suggesting some core limitations in the models.
+Future work could look at the effect of different hyperparameters, model architectures, and data modifications to improve the prediction capabilities of the models.
 

### 2024-03-29 paper/paper.tex
more paper stuff

@@
 \begin{table}[!h]
-    % ref for this table
-
     \centering
@@
 % model x had the best loss in the end 
-Looking at the results in \ref{fig:different_backbones} of the inital choice of hyperparameters for the different models the first thing we noticed was similar behavior of all the $R^2$ values in terms of their sign.
+Looking at the results in \ref{fig:different_backbones} of the initial choice of hyperparameters for the different models the first thing we noticed was similar behavior of all the $R^2$ values in terms of their sign.
 The coefficient of determination is defined as follows
@@
     \includegraphics[width=1\textwidth]{assets/different_backbones.png}
-    \caption{Validation loss for primary task with different backbones}
+    \caption{Validation $R^2$ for primary task with different backbones}
     \label{fig:different_backbones}
@@
 
-While both of the EfficientNet backbones exhibit a similar behavior as they come from the same family of models; this being a fast convergence though also an early plateu at a more suboptimal loss.
-The ResNet50 backbone on the other hand has a slower convergence, but a more consisestent decrease in loss (increase in $R^2$).
+While both of the EfficientNet backbones exhibit similar behavior as they come from the same family of models; this being a fast convergence though also an early plateau at a more suboptimal loss.
+The ResNet50 backbone on the other hand has a slower convergence, but a more consistent decrease in loss (increase in $R^2$).
 This could indicate that the ResNet50 backbone has a greater potential to learn from the data albeit in a slower manner.
@@
     \centering
-    \includegraphics[width=0.4\textwidth]{assets/lr_schedule_diffs.png}
+    \includegraphics[width=1\textwidth]{assets/lr_schedule_diffs.png}
     \caption{Comparison of different learning rate schedules}
@@
 We wanted to test out whether there was any difference in $R^2$ whether we allowed the model to adapt the weights of the backbone or not.
-The results of this can be seen in \ref{fig:backbone_training} in both of these instances we see the model converging at around the 2nd epoch and hitting a plateu of $\approx -0.03 R^2$.
-This is likely due to the fact that that the backbone requires little additional training to learn the features from the data and generalizes well to this task.
+The results of this can be seen in \ref{fig:backbone_training} in both of these instances we see the model converging at around the 2nd epoch and hitting a plateau of $\approx -0.03 R^2$.
+This is likely due to the fact that the backbone requires little additional training to learn the features from the data and generalizes well to this task.
 
@@
 
-To construct the CNN trained on images and geodata we first created the two branches of the model, one to take in the image data and reduce it to a 1-tensor and similarly one to extract features from the tabular data.
-The two branches where then concatenated and fed to the main and auixallary task dense output layers.
+To construct the CNN trained on images and geodata we first created the two branches of the model, one to take in the image data and reduce it to a 1-tensor and similarly, one to extract features from the tabular data.
+The two branches were then concatenated and fed to the main and auxiliary task dense output layers.
 As the only difference here is the additional branch for the geodata we kept the hyperparameters the same to get a fair comparison between both approaches.
@@
 The results of this can be seen in \ref{fig:basic_evaluation}.
-We can see that the model trained on images and geodata has a higher $R^2$ suggesting a relative improvement in the prediction capability over just the image model which has a best recorded $R^2$ of <number>.
+We can see that the model trained on images and geodata has a lower $R^2$ than the model just trained on images, though not by any significant margin.
 
@@
     \centering
-    \includegraphics[width=0.7\textwidth]{assets/fin_img_vs_geo.png}
+    \includegraphics[width=0.8\textwidth]{assets/fin_img_vs_geo.png}
     \caption{Basic evaluation metrics of the two models}
@@
 
-% give a simple summary of the results
-One potential reason why the combined model performs better is simply that certain patterns are easier to learn from the geodata than from the images.
-Take the trait \textit{Stem specific density} for example, this is a trait that is likely to be more defined as a function of the environment \cite{plants8030065} rather than any visual cues.
-The same applies to the trait \textit{Leaf nitrogen content} which has been shown to be associated with abiotic stress (e.g.
-heat, cold, drought, etc.) \cite{Ye2022} and thus is likely to be more easily learned from the geodata.
+Even though the $R^2$ did not vary by a significant margin between the combined and image model, the $R^2$ on the test set did.
+While both models still had a negative $R^2$ the test set the image model had a higher $R^2$ than the combined model.
+One potential explanation for this is that the model is overfitting on some property derived from the training data.
+The source of this issue would then reasonably lie with the inclusion of the geodata.
+While in theory the geodata does provide more context to the model and thus should give it a greater potential to learn from the data, it could also be the case that the model is learning from the geodata in a way that is not beneficial to the prediction of the plant traits.
+
+\begin{table}[!h]
+    \centering
+    \begin{tabular}{@{}llll@{}}
+    \toprule
+    & Image model & Combined model \\ \midrule
+    test set $R^2$              & -6.18163             & -381.62190   \\ \bottomrule
+\end{tabular}
+\caption{Comparison of the $R^2$ on the test set for the two models}
+\end{table}
 
@@
 % discuss the results
-The prediction of plant traits from images and geodata is a complex task that requires a CNN to learn and extract featues from both images and geodata.
-The results of our work show that a CNN trained on just image outpreforms one trained on images and geodata.
-Though neither of the models elaborated here is necessarily a good predictior, one of the main reasons for this is likely the complexity and nuance of the task.
+The prediction of plant traits from images and geodata is a complex task that requires a CNN to learn and extract features from both images and geodata.
+The results of our work show that a CNN trained on just an image outperforms one trained on images and geodata.
+Though neither of the models elaborated here is necessarily a good predictor, one of the main reasons for this is likely the complexity and nuance of the task.
 We now propose various directions for future work to improve the prediction capabilities of the models.
@@
 Hyperparameter tuning is one of the most important aspects of training a machine learning model.
-In our work we only tested a few different learning rate schedules and backbones.
-Random search - which as been demonstrated to be more effective than grid search \cite{JMLR:v13:bergstra12a} - would likely have been a better choice, albeit more computationally expensive.
-In addition we predominantly looked at hyperparameters, but varying the model architecture could have also been a benifical direction.
+In our work, we only tested a few different learning rate schedules and backbones using a greedy optimization approach.
+Random search - which has been demonstrated to be more effective than grid search \cite{JMLR:v13:bergstra12a} - would likely have been a better choice, albeit more computationally expensive.
+In addition, we predominantly looked at hyperparameters, but varying the model architecture could have also been a beneficial direction.
 A larger model could potentially learn more nuanced features in the data.
@@
 Certain plant traits rely on some sense of relative scale to be elicited accurately (e.g.
-we know that a tree is bigger than flower regardless of image angle), this scale is easily lost due to the variability in how people take photos.
+we know that a tree is bigger than a flower regardless of image angle), but this scale is easily lost due to the variability in how people take photos.
 One way of mitigating this could be to incorporate scale normalization through Depth Estimation, previous work \cite{Ummenhofer_2017} demonstrated the ability to train CNNs on the DIODE (Dense Indoor/Outdoor DEpth) \cite{vasiljevic2019diode} dataset to estimate depth.
@@
 
-How adaptable pretrained models are to new tasks is also is also something that could be further explored.
+How adaptable pre-trained models are to new tasks is also is also something that could be further explored.
 Kunze et al.
@@
 This reasoning could possibly be applied to more accurately find the optimal backbone.
-Additionally the use of a pretrained model could be further explored by looking at the effect of the number of layers frozen during training.
+Additionally, the use of a pre-trained model could be further explored by looking at the effect of the number of layers frozen during training.
 
 \section{Conclusion}
-To conclude we noticed that first and foremost the task to predict pant traits is a hard one.
-It requires a careful consideration during every step of the process from understanding the data to how a neural network architecture will learn from it.
-From our results we observed that a model trained on images and ancillary geodata converges faster but does not necessarily have a higher $R^2$ than a model trained on images alone.
+To conclude we noticed that first and foremost the task of predicting plant traits is a hard one.
+It requires careful consideration during every step of the process from understanding the data to how a neural network architecture will learn from it.
+From our results, we observed that a model trained on images and ancillary geodata converges faster but does not necessarily have a higher $R^2$ than a model trained on images alone.
 In the final tests run these results were also observed with our image model having a higher $R^2$ than the combined model.
-Though both models still have a negative $R^2$ suggesting some core limitations in the models.
+However, both models still have a negative $R^2$ suggesting some core limitations in the models.
 Future work could look at the effect of different hyperparameters, model architectures, and data modifications to improve the prediction capabilities of the models.

### 2024-03-29 paper/paper.tex
add abstract

@@
     % Abstract content goes here
+    Global warming and climate change have created concerns about the adaptation of plant species to constantly evolving environmental conditions.
+Monitoring and analyzing plant traits provides crucial insights into ecosystem responses to such changes.
+Convolutional Neural Networks (CNNs) offer a promising approach to predicting plant traits based on images, and related geological data.
+This paper explores the effectiveness of CNNs predicting relevant plant traits trained on images alone versus those trained on images combined with ancillary geodata.
+Our results indicate that while models trained on images and geodata converge faster, they do not necessarily outperform models trained on images alone.
+Further investigation into hyperparameters, model architectures, and data modifications is warranted to enhance the prediction capabilities of these models and understanding the relationship between plant traits and environmental factors.
 \end{abstract}
@@
 \cite{pmlr-v80-chen18a} which demonstrated that for very similar or mostly similar auxiliary tasks the contribution is generally a positive one.
-Based on this we decided to use a simple k-nearest neighbor (kNN) imputation strategy which has generally proved effective \cite{joel2024performance}.
+Based on this we decided to use a simple k-nearest neighbor (kNN) imputation strategy which has generally proven to be effective \cite{joel2024performance}.
 The rest of the data had all values present.

### 2024-03-29 paper/paper.tex
spelling

--- 2024-03-29 paper/paper.tex
@@
 Integral to any machine learning model is the data used to train, validate, and ultimately test the model.
-The data used consisted of 3 main components: ancillary geodata from various sources, the main task trait means and auxiliary task standard deviations, and the training ima ges of the plants.
+The data used consisted of 3 main components: ancillary geodata from various sources, the main task trait means and auxiliary task standard deviations, and the training images of the plants.
 
 \smallskip 
-Upon visual inspection one of the first issues we spotted was a considerable chunk (29.53\%) of data missing for the auxiliary task standard deviations.
+Upon visual inspection, one of the first issues we spotted was a considerable chunk (29.53\%) of data missing for the auxiliary task standard deviations.
 Through some previous work on the same Kaggle, we noticed that the auxiliary data was a useful inclusion.
@@
 We are using \textit{efficientnetv2\_b2\_imagenet} and \textit{efficientnetv2\_s\_imagenet} which are trained on imagenet, with the ladder being a smaller (fewer parameters) version of the former.
-Comparing these backbones in a table we can get some overview of how they stack up as seen in \ref{tab:backbone_comparison}
+Comparing these backbones in a table, we can get some overview of how they stack up, as seen in \ref{tab:backbone_comparison}
 
-To compare these backbones we trained 3 different models with the same architecture but different backbones.
+To compare these backbones, we trained 3 different models with the same architecture but different backbones.
 We also specified a contribution of 0.3 to the final loss of the auxiliary task (as opposed to 1.0 for the main task) as to limit its impacts on the direction of training.
-A visual overview of the different models can be seen in \ref{fig:models_overview}.
+A visual overview of the different models, can be seen in \ref{fig:models_overview}.
 
@@
 This is a clear indication that the model is fitting against the data, numerious reasons could be the cause of this.
-Due to the results of the intial training not being conclusive in regards to which model is better we decided to to simply continue with the model using the \textit{efficientnetv2\_b2\_imagenet} backbone as it has a high top 5 accuracy while also having fewer parameters.
+Due to the results of the intial training not being conclusive in regards to which model is better, we decided to simply continue with the model using the \textit{efficientnetv2\_b2\_imagenet} backbone as it has a high top 5 accuracy while also having fewer parameters.
 
@@
 
-While both of the EfficientNet backbones exhibit similar behavior as they come from the same family of models; this being a fast convergence though also an early plateau at a more suboptimal loss.
+Both of the EfficientNet backbones exhibit similar behavior as they come from the same family of models; this being a fast convergence though also an early plateau at a more suboptimal loss.
 The ResNet50 backbone on the other hand has a slower convergence, but a more consistent decrease in loss (increase in $R^2$).
@@
 
-To construct the CNN trained on images and geodata we first created the two branches of the model, one to take in the image data and reduce it to a 1-tensor and similarly, one to extract features from the tabular data.
+To construct the CNN trained on images and geodata, we first created the two branches of the model, one to take in the image data and reduce it to a 1-tensor and similarly, one to extract features from the tabular data.
 The two branches were then concatenated and fed to the main and auxiliary task dense output layers.
-As the only difference here is the additional branch for the geodata we kept the hyperparameters the same to get a fair comparison between both approaches.
+As the only difference here is the additional branch for the geodata, we kept the hyperparameters the same to get a fair comparison between both approaches.
 A visual overview of the model can be seen in \ref{fig:geodata_model}.
@@
 
-\smallskip
-
 % comparative evaluation
-
-\section{Results}
-
-Even though the $R^2$ did not vary by a significant margin between the combined and image model, the $R^2$ on the test set did.
-While both models still had a negative $R^2$ the test set the image model had a higher $R^2$ than the combined model.
-One potential explanation for this is that the model is overfitting on some property derived from the training data.
-The source of this issue would then reasonably lie with the inclusion of the geodata.
-While in theory the geodata does provide more context to the model and thus should give it a greater potential to learn from the data, it could also be the case that the model is learning from the geodata in a way that is not beneficial to the prediction of the plant traits.
-
 \begin{table}[!h]
@@
 
+\section{Results}
+
+Even though the $R^2$ did not vary by a significant margin between the combined and image model, the $R^2$ on the test set did.
+While both models still had a negative $R^2$ for the test set the image model had a higher $R^2$ than the combined model.
+One potential explanation for this is that the model is overfitting on some property derived from the training data.
+The source of this issue would then reasonably lie with the inclusion of the geodata.
+While in theory the geodata does provide more context to the model and thus should give it a greater potential to learn from the data, it could also be the case that the model is learning from the geodata in a way that is not beneficial to the prediction of the plant traits.
+
 \section{Discussion}
@@
 The results of our work show that a CNN trained on just an image outperforms one trained on images and geodata.
-Though neither of the models elaborated here is necessarily a good predictor, one of the main reasons for this is likely the complexity and nuance of the task.
+Though neither of the models elaborated here are necessarily good predictors, one of the main reasons for this is likely the complexity and nuance of the task.
 We now propose various directions for future work to improve the prediction capabilities of the models.

### 2024-03-29 paper/paper.tex
info sheet update

@@
 The results of our work show that a CNN trained on just an image outperforms one trained on images and geodata.
-Though neither of the models elaborated here are necessarily good predictors, one of the main reasons for this is likely the complexity and nuance of the task.
+Neither of the models elaborated here are necessarily good predictors, one of the main reasons for this is likely the complexity and nuance of the task regression task.
 We now propose various directions for future work to improve the prediction capabilities of the models.

### 2024-03-29 paper/info_sheet.tex

@@
 
-\paragraph{Software used} \textit{Describe briefly which software and libraries you relied on. If you built a particular algorithm from scratch, you can explain that here.}
+\paragraph{Software used} \textit{jupyter, python3.10, keras3, tf2.16, various data viz libraries, gcp compute engine to train the models, pandas for most of the data science related tasks}
 

### 2024-03-29 paper/paper.tex

@@
 
-How adaptable pre-trained models are to new tasks is also is also something that could be further explored.
+How adaptable pre-trained models are to new tasks is also something that could be further explored.
 Kunze et al.
@@
 \section{Conclusion}
-To conclude we noticed that first and foremost the task of predicting plant traits is a hard one.
+To conclude, we noticed that first and foremost the task of predicting plant traits is a hard one.
 It requires careful consideration during every step of the process from understanding the data to how a neural network architecture will learn from it.
 From our results, we observed that a model trained on images and ancillary geodata converges faster but does not necessarily have a higher $R^2$ than a model trained on images alone.
-In the final tests run these results were also observed with our image model having a higher $R^2$ than the combined model.
+In the final test run these results were also observed with our image model having a higher $R^2$ than the combined model.
 However, both models still have a negative $R^2$ suggesting some core limitations in the models.

### 2024-03-29 paper/paper.tex
added abstract from red branch

@@
 \begin{abstract}
-    % Abstract content goes here
+    Global warming and climate change have created concerns about the adaptation of plant species to constantly evolving environmental conditions.
+Monitoring and analyzing plant traits provides crucial insights into ecosystem responses to such changes.
+Convolutional Neural Networks (CNNs) offer a promising approach to predicting plant traits based on images, and related geological and satilite data.
+This paper explores the effectiveness of CNNs predicting relevant plant traits trained on images alone versus those trained on images combined with ancillary geodata.
+Our results indicate that while models trained on images and geodata converge faster, they do not necessarily outperform models trained on images alone.
+Further investigation into hyperparameters, model architectures, and data modifications is warranted to enhance the prediction capabilities of these models and understanding the relationship between plant traits and environmental factors.
 \end{abstract}

### 2024-03-29 paper/paper.tex
fixes

@@
     \includegraphics[width=1\textwidth]{assets/lr_schedule_diffs.png}
-    \caption{Comparison of different learning rate schedules}
+    \caption{Comparison of different learning rate schedules ; (left) step-based ; (center) cosine annealing ; (right) exponential}
     \label{fig:lr_schedules}
@@
 Another point of consideration we had was the training of the backbone.
-We wanted to test out whether there was any difference in $R^2$ whether we allowed the model to adapt the weights of the backbone or not.
+We wanted to test out whether there was any difference in $R^2$ when we allowed the model to adapt the weights of the backbone or not.
 The results of this can be seen in \ref{fig:backbone_training} in both of these instances we see the model converging at around the 2nd epoch and hitting a plateau of $\approx -0.03 R^2$.

### 2024-03-29 paper/paper.tex
fixes
conclusion and fixed comma's, grammar and spelling

@@
 Monitoring and analyzing plant traits provides crucial insights into ecosystem responses to such changes.
-Convolutional Neural Networks (CNNs) offer a promising approach to predicting plant traits based on images, and related geological and satilite data.
+Convolutional Neural Networks (CNNs) offer a promising approach to predicting plant traits based on images and related geological and satellite data.
 This paper explores the effectiveness of CNNs predicting relevant plant traits trained on images alone versus those trained on images combined with ancillary geodata.
 Our results indicate that while models trained on images and geodata converge faster, they do not necessarily outperform models trained on images alone.
-Further investigation into hyperparameters, model architectures, and data modifications is warranted to enhance the prediction capabilities of these models and understanding the relationship between plant traits and environmental factors.
+Further investigation into hyperparameters, model architectures, and data modifications is warranted to enhance these models' predictive capabilities and understand the relationship between plant traits and environmental factors.
 \end{abstract}
@@
 % primary paragraph 
-Global warming and more broadly climate change have become a major concern for the world as its effects are becoming more and more apparent \cite{WANG2023100237}.
-Changing weather patterns, especially towards more extreme conditions are causing plants to adapt to new environments \cite{GRAY201664}.
+Global warming and, more broadly, climate change have become a major concern for the world as its effects are becoming increasingly apparent \cite{WANG2023100237}.
+Changing weather patterns, especially towards more extreme conditions, are causing plants to adapt to new environments \cite{GRAY201664}.
 One method of measuring said adaptation is to look at the traits, that is, properties of a plant that describe how it functions and interacts with the environment.
-These traits include but are not limited to the plant height, leaf area, but also dry mass, and leaf nitrogen content amongst various others.
+These traits include but are not limited to the plant height, leaf area, dry mass, and leaf nitrogen content, amongst various others.
 Monitoring these traits allows us to gain vital insights into how climate change impacts different ecosystems.
-While somewhat simple manual measurement techniques exist, at scale they are not feasible.
+While somewhat simple manual measurement techniques exist, they are not feasible at scale.
 This is where Convolutional Neural Networks (CNN) come in.
@@
 \smallskip
-Through the work demonstrated by Schiller et al \cite{schiller2021deep} we know that CNNs can be used to predict plant traits from images.
-The images used to train this network came from \textit{citizen science photographs} which are images taken by citizens of plants from all across the world using AI plant species identification apps (e.g.
-iNaturalist, Pl@ntNet).
+Through the work demonstrated by Schiller et al.
+\cite{schiller2021deep}, we know that CNNs can be used to predict plant traits from images.
+The images used to train this network came from \textit{citizen science photographs} and are taken by citizens of plants worldwide using AI plant identification apps (e.g., iNaturalist, Pl@ntNet).
 Citizen science photographs also come with location metadata, which can be used to extract ancillary geodata such as precipitation, temperature, and soil type.
-This geodata can optionally be combined with the images to create a CNN which can potentially learn to extract features from images in conjunction with geodata to predict plant traits.
+This geodata can optionally be combined with the images to create a CNN, which can potentially learn to extract features from images in conjunction with geodata to predict plant traits.
 
 \smallskip
-For our method, we wanted to compare the accuracy of a CNN trained on images alone and using a pre-trained backbone (e.g.
-ResNet, EfficientNet, etc.) to a CNN trained on images and geodata.
+For our method, we wanted to compare the accuracy of a CNN trained on images alone and using a pre-trained backbone (e.g., ResNet, EfficientNet, etc.) to a CNN trained on images and geodata.
 We hypothesize that the CNN trained on images and geodata will outperform the CNN trained on images alone.
-With our results we hope to provide a better understanding of the key factors that are needed to predict plant traits.
+With our results, we hope to provide a better understanding of the key factors needed to predict plant traits.
 
@@
 
-Integral to any machine learning model is the data used to train, validate, and ultimately test the model.
-The data used consisted of 3 main components: ancillary geodata from various sources, the main task trait means and auxiliary task standard deviations, and the training images of the plants.
+Integral to any machine learning model is the data used to train, validate, and ultimately test it.
+The data used consisted of three main components: ancillary geodata from various sources, the main task trait means and auxiliary task standard deviations, and the training images of the plants.
 
@@
 Upon visual inspection, one of the first issues we spotted was a considerable chunk (29.53\%) of data missing for the auxiliary task standard deviations.
-Through some previous work on the same Kaggle, we noticed that the auxiliary data was a useful inclusion.
-We can also validate these reports with previous work applying auxiliary tasks in CNNs such as the work of Lukas Libel and Marco Kroener which found that the inclusion of an auxiliary task with minor relevance to the main task did indeed boost performance \cite{lukaslibel} and Chen et.
-al.
-\cite{pmlr-v80-chen18a} which demonstrated that for very similar or mostly similar auxiliary tasks the contribution is generally a positive one.
-Based on this we decided to use a simple k-nearest neighbor (kNN) imputation strategy which has generally proven to be effective \cite{joel2024performance}.
+Previous work on the same Kaggle showed us that the auxiliary data was a useful inclusion.
+We can also validate these reports with previous work applying auxiliary tasks in CNNs, such as the work of Lukas Libel and Marco Kroener, which found that the inclusion of an auxiliary task with minor relevance to the main task did indeed boost performance \cite{lukaslibel}.
+Chen et al.
+\cite{pmlr-v80-chen18a} demonstrated that for very similar or mostly similar auxiliary tasks, the contribution is generally a positive one.
+Based on this, we decided to use a simple k-nearest neighbor (kNN) imputation strategy, which has generally proven effective \cite{joel2024performance}.
 The rest of the data had all values present.
@@
 The geodata was another point of consideration in the data pre-processing phase.
-Looking at the individual columns we noticed that the instances had features that encoded similar information and thus were likely redundant in the training process.
-Since we are working with a very high dimensional feature space to get some sort of visual verification of this claim we plotted a correlation heatmap for the 3 biggest groups of the geodata, namely the datasets: SOIL, MODIS, and VOD.
+Looking at the individual columns, we noticed that the instances had features that encoded similar information and, thus, were likely redundant during the training process.
+Since we are working with a very high dimensional feature space to get some visual verification of this claim, we plotted a correlation heatmap for the three biggest groups of the geodata, namely the datasets: SOIL, MODIS, and VOD.
 
@@
 
-Since these datasets individually can be broken down into groups of features generally talking about a similar topic (e.g.
-precipitation metrics, temperature metrics, etc) we can see that the correlation within these groups is quite high.
+Since these datasets can be broken down into groups of features that generally talk about a similar topic (e.g., precipitation metrics, temperature metrics, etc.), we can see that the correlation within these groups is relatively high.
 Even though previous work \cite{DBLP:journals/corr/abs-2007-00062} has demonstrated the viability of neural network models to learn from high dimensional data, removing redundant features has in numerous instances been shown to improve model performance \cite{chen2022survey}.
@@
 % talk about the augmentation 
-Our training and test image instance - consisting of science citizen photographs - were taken by a wide variety of people under a wide variety of conditions.
-This means that the images are of varying quality, resolution, and orientation.
+Our training and test image instance - consisting of science citizen photographs - was taken by a wide variety of people under a wide variety of conditions.
+This means the images are of varying quality, resolution, and orientation.
 We can see this just by taking a random sample of 5 images.
@@
 
-To prevent our model from fitting to noise in the data as a byproduct of how images were taken we decided to apply different keras preprocessing layers to batches concurrently.
-A key benefit of this would be that the model can generalize better to unseen data by learning potentially more meaningful features.
-While this approach of applying augmentations in the preprocessing stage is less flexible than having filters integrated into the model it does reduce model complexity though at the cost of potentially losing some ability to learn more complex features by being able to dynamically adjust the filters.
-Sampling again we can look at the results of our augmentations.
+To prevent our model from fitting to noise in the data as a byproduct of how images were taken, we applied different Keras preprocessing layers to batches concurrently.
+A key benefit would be that the model can generalize better to unseen data by learning potentially more meaningful features.
+While this approach of applying augmentations in the preprocessing stage is less flexible than having filters integrated into the model, it does reduce model complexity, though at the cost of potentially losing some ability to learn more complex features by being able to adjust the filters dynamically.
+After sampling again, we can look at the results of our augmentations.
 
@@
 \smallskip 
-In addition to augmentation we noted that - though not too common - there was still a presence of certain substantial outliers in our data.
-Through visual inspection a common reason seems to have been specifically how close an indivdiual was when they took an image.
-As the training data itself was labelled by a machine learning model in a supervised fashion its evident perspective was not something the intial model had learned.
-As this scaling issue is a byproduct of the inital model which labelled the data and not any true property of the data we decided to remove these outliers.
-We did this by constraining the target distributions to the interval $[0.0001, 0.9999]$ and then removing any instances that were outside of this interval.
+In addition to augmentation, we noted that - though not too common - there was still a presence of certain substantial outliers in our data.
+Through visual inspection, a common reason seems to have been specifically how close an individual was when they took an image.
+As the training data was labeled by a machine learning model in a supervised fashion, its evident perspective was not something the initial model had learned.
+As this scaling issue is a byproduct of the initial model, which labeled the data and not any true property of the data, we decided to remove these outliers.
+We did this by constraining the target distributions to the interval $[0.0001, 0.9999]$ and removing any instances outside of this interval.
 
 \smallskip
-Another method of reducing overfitting we thought relevant to apply here would be k-fold cross-validation for the data split.
-We chose $k=5$ as for large datasets this has proven effective \cite{Marcot2021} and is generally around the commonly chosen value of 10 \cite{kequal10isgood}, where the 0th fold was designated as the validation set.
-Given that the data is from a geographically large enough sample we decided to create the folds in a stratified manner under the assumption the distribution of traits would be reasonably representative of the true distribution and thus one the model should learn.
-This is important as it ensures that the model is trained on a representative sample of the data in each fold and in turn, has its predictions distributed in a similar manner to the assumed true distribution of the traits.
-So we reduce the risk of overfitting to the training data while also ensuring the model is trained using a representative sample of targets in each fold.
-To validate that the distributions of the target traits are indeed similar we averaged the distributions of the target traits in each fold overlayed this on the original distribution of trait means and normalized them to the same scale.
+Another method of reducing overfitting relevant to apply here is k-fold cross-validation for the data split.
+We chose $k=5$, as for large datasets, this has proven effective \cite{Marcot2021} and is generally around the commonly chosen value of 10 \cite{kequal10isgood}, where the 0th fold was designated as the validation set.
+Given that the data is from a geographically large enough sample, we decided to create the folds in a stratified manner under the assumption that the distribution of traits would be reasonably representative of the true distribution and, thus, one the model should learn.
+This is important as it ensures that the model is trained on a representative sample of the data in each fold and, in turn, has its predictions distributed similarly to the assumed true distribution of the traits.
+So, we reduce the risk of overfitting to the training data while also ensuring the model is trained using a representative sample of targets in each fold.
+To validate that the distributions of the target traits are indeed similar, we averaged the distributions of the target traits in each fold, overlayed this on the original distribution of trait means, and normalized them to the same scale.
 
@@
 The model consists of a trained backbone; in our instance, we compared \textit{efficientnetv2\_b2\_imagenet}, \textit{efficientnetv2\_s\_imagenet}, and \textit{resnet50\_imagenet}.
-The backbone outputs then feed into a global average pooling layer, followed by 2 dense layers and a dropout layer connected to the two dense output layers for the main and auxiliary tasks.
-The hidden layers use a ReLU activation function, the outputs use no activation function as the targets are continuous.
+The backbone outputs then feed into a global average pooling layer, followed by two dense layers and a dropout layer connected to the two dense output layers for the main and auxiliary tasks.
+The hidden layers use a ReLU activation function, and the outputs use no activation function as the targets are continuous.
 The model is trained using the Adam optimizer with a learning rate of 1e-4, a batch size of 48, and for 12 epochs.
-Additionally, we used the $R^2$ metric for both loss and evaluation of the model performance as this is a regression problem.
-Finally, we used step-based for the learning rate schedule which was defined as follows
+Additionally, we used the $R^2$ metric for both loss and evaluation of the model performance, as this is a regression problem.
+Finally, we used step-based for the learning rate schedule, which was defined as follows:
 \[
@@
 \smallskip
-Residual Network (ResNet) \cite{he2016identity} introduced in 2016 is a deep learning model that is able to learn features from images.
-Most famously it's known for mitigating the vanishing gradient problem through the introduction of identity skip connections.
-We are using \text{ResNet50} which was trained on imagenet and consists of 48 convolution layers, in addition to a MaxPool and AveragePool layer.
-EfficientNetV2 \cite{tan2021efficientnetv2} is a family of CNNs from 2021 that have faster speeds than many previous models primarily by increasing image size during training and incorporating adaptive regularization to compensate for any potential accuracy loss.
-We are using \textit{efficientnetv2\_b2\_imagenet} and \textit{efficientnetv2\_s\_imagenet} which are trained on imagenet, with the ladder being a smaller (fewer parameters) version of the former.
+Residual Network (ResNet) \cite{he2016identity} introduced in 2016 is a deep learning model that can learn features from images.
+Most famously, it is known for mitigating the vanishing gradient problem by introducing identity skip connections.
+We are using \text{ResNet50}, which was trained on imagenet and consists of 48 convolution layers, in addition to a MaxPool and AveragePool layer.
+EfficientNetV2 \cite{tan2021efficientnetv2} is a family of CNNs from 2021 that have faster speeds than many previous models, primarily by increasing image size during training and incorporating adaptive regularization to compensate for any potential accuracy loss.
+We are using \textit{efficientnetv2\_b2\_imagenet} and \textit{efficientnetv2\_s\_imagenet}, which are trained on imagenet, with the latter being a smaller (fewer parameters) version of the former.
 Comparing these backbones in a table, we can get some overview of how they stack up, as seen in \ref{tab:backbone_comparison}
 
-To compare these backbones, we trained 3 different models with the same architecture but different backbones.
-We also specified a contribution of 0.3 to the final loss of the auxiliary task (as opposed to 1.0 for the main task) as to limit its impacts on the direction of training.
-A visual overview of the different models, can be seen in \ref{fig:models_overview}.
+We trained three models with the same architecture but different backbones to compare these backbones.
+We also specified a contribution of 0.3 to the final loss of the auxiliary task (as opposed to 1.0 for the main task) to limit its impacts on the direction of training.
+A visual overview of the different models can be seen in \ref{fig:models_overview}.
 
@@
 % model x had the best loss in the end 
-Looking at the results in \ref{fig:different_backbones} of the initial choice of hyperparameters for the different models the first thing we noticed was similar behavior of all the $R^2$ values in terms of their sign.
-The coefficient of determination is defined as follows
+Looking at the results in \ref{fig:different_backbones} of the initial choice of hyperparameters for the different models, we first noticed similar behavior of all the $R^2$ values in terms of their sign.
+The coefficient of determination is defined as follows:
 
@@
 Where $SS_{res}$ is the residual sum of squares and $SS_{tot}$ is the total sum of squares.
-One observation in our results was that for all instances we had a negative $R^2$, implying that $SS_{res} > SS_{tot}$ meaning our model preforms worse than the baseline at $SS_{res} = SS_{tot}$.
-This is a clear indication that the model is fitting against the data, numerious reasons could be the cause of this.
-Due to the results of the intial training not being conclusive in regards to which model is better, we decided to simply continue with the model using the \textit{efficientnetv2\_b2\_imagenet} backbone as it has a high top 5 accuracy while also having fewer parameters.
+One observation in our results was that for all instances, we had a negative $R^2$, implying that $SS_{res} > SS_{tot}$ meaning our model performs worse than the baseline at $SS_{res} = SS_{tot}$.
+This is a clear indication that the model is fitting against the data.
+Numerous reasons could be the cause of this.
+Due to the results of the initial training not being conclusive in regards to which model is better, we decided to simply continue with the model using the \textit{efficientnetv2\_b2\_imagenet} backbone as it has a high top 5 accuracy while also having fewer parameters.
 
@@
 
-Both of the EfficientNet backbones exhibit similar behavior as they come from the same family of models; this being a fast convergence though also an early plateau at a more suboptimal loss.
-The ResNet50 backbone on the other hand has a slower convergence, but a more consistent decrease in loss (increase in $R^2$).
-This could indicate that the ResNet50 backbone has a greater potential to learn from the data albeit in a slower manner.
+Both of the EfficientNet backbones exhibit similar behavior as they come from the same family of models; this being a fast convergence, though also an early plateau at a more suboptimal loss.
+The ResNet50 backbone, on the other hand, has a slower convergence but a more consistent decrease in loss (increase in $R^2$).
+This could indicate that the ResNet50 backbone has a greater potential to learn from the data, albeit in a slower manner.
 
 \smallskip 
-One of the first things we decided to test was the learning rate schedule.
-We tested out 3 different schedules: step-based (baseline), exponential, and cosine annealing.
+The learning rate schedule was one of the first things we decided to test.
+We tested three schedules: step-based (baseline), exponential, and cosine annealing.
 The results demonstrated in \ref{fig:lr_schedules} show that the cosine annealing and exponential schedule had a slower convergence than the step-based schedule and a more erratic behavior.
@@
 Another point of consideration we had was the training of the backbone.
-We wanted to test out whether there was any difference in $R^2$ when we allowed the model to adapt the weights of the backbone or not.
-The results of this can be seen in \ref{fig:backbone_training} in both of these instances we see the model converging at around the 2nd epoch and hitting a plateau of $\approx -0.03 R^2$.
+We wanted to test whether there was any difference in $R^2$ when we allowed the model to adapt the weights of the backbone.
+These results can be seen in \ref{fig:backbone_training}.
+In both instances, we see the model converging at around the second epoch and hitting a plateau of $\approx -0.03 R^2$.
 This is likely due to the fact that the backbone requires little additional training to learn the features from the data and generalizes well to this task.
@@
 
-To construct the CNN trained on images and geodata, we first created the two branches of the model, one to take in the image data and reduce it to a 1-tensor and similarly, one to extract features from the tabular data.
+To construct the CNN trained on images and geodata, we first created the two branches of the model, one to take in the image data and reduce it to a 1-tensor and, similarly, one to extract features from the tabular data.
 The two branches were then concatenated and fed to the main and auxiliary task dense output layers.
@@
 
-For the final comparison we first looked at the basic evaluation metrics of either model.
-The results of this can be seen in \ref{fig:basic_evaluation}.
+For the final comparison, we first looked at the basic evaluation metrics of either model.
+These results can be seen in \ref{fig:basic_evaluation}.
 We can see that the model trained on images and geodata has a lower $R^2$ than the model just trained on images, though not by any significant margin.
@@
 Even though the $R^2$ did not vary by a significant margin between the combined and image model, the $R^2$ on the test set did.
-While both models still had a negative $R^2$ for the test set the image model had a higher $R^2$ than the combined model.
-One potential explanation for this is that the model is overfitting on some property derived from the training data.
+While both models still had a negative $R^2$ for the test set, the image model had a higher $R^2$ than the combined model.
+One potential explanation is that the model is overfitting on some property derived from the training data.
 The source of this issue would then reasonably lie with the inclusion of the geodata.
-While in theory the geodata does provide more context to the model and thus should give it a greater potential to learn from the data, it could also be the case that the model is learning from the geodata in a way that is not beneficial to the prediction of the plant traits.
+While, in theory, the geodata does provide more context to the model and thus should give it a greater potential to learn from the data, it could also be the case that the model is learning from the geodata in a way that is not beneficial to the prediction of the plant traits.
 
@@
 % discuss the results
-The prediction of plant traits from images and geodata is a complex task that requires a CNN to learn and extract features from both images and geodata.
+Predicting plant traits from images and geodata is a complex task requiring a CNN to learn and extract features from images and geodata.
 The results of our work show that a CNN trained on just an image outperforms one trained on images and geodata.
-Neither of the models elaborated here are necessarily good predictors, one of the main reasons for this is likely the complexity and nuance of the task regression task.
-We now propose various directions for future work to improve the prediction capabilities of the models.
+Neither of the models elaborated here is necessarily a good predictor; one of the main reasons for this is likely the complexity and nuance of the regression task.
+We now propose various directions for future work to improve the models' prediction capabilities.
 
@@
 In addition, we predominantly looked at hyperparameters, but varying the model architecture could have also been a beneficial direction.
-A larger model could potentially learn more nuanced features in the data.
+A larger model could learn more nuanced features in the data.
 
-The training data itself can also be possibly modified to improve the model.
-Certain plant traits rely on some sense of relative scale to be elicited accurately (e.g.
-we know that a tree is bigger than a flower regardless of image angle), but this scale is easily lost due to the variability in how people take photos.
-One way of mitigating this could be to incorporate scale normalization through Depth Estimation, previous work \cite{Ummenhofer_2017} demonstrated the ability to train CNNs on the DIODE (Dense Indoor/Outdoor DEpth) \cite{vasiljevic2019diode} dataset to estimate depth.
-The learned depth information of the images could be used then as a moderator to scale the images to a common scale possibly improving the prediction capabilities.
+The training data itself can also be modified to improve the model.
+Certain plant traits rely on some sense of relative scale to be elicited accurately (e.g., we know that a tree is bigger than a flower regardless of image angle).
+However, this scale is easily lost due to the variability in how people take photos.
+One way of mitigating this could be to incorporate scale normalization through Depth Estimation; previous work \cite{Ummenhofer_2017} demonstrated the ability to train CNNs on the DIODE (Dense Indoor/Outdoor DEpth) \cite{vasiljevic2019diode} dataset to estimate depth.
+The images' learned depth information could then be used as a moderator to scale the images to a common scale, possibly improving the prediction capabilities.
 
-How adaptable pre-trained models are to new tasks is also something that could be further explored.
+How adaptable pre-trained models are to new tasks could also be further explored.
 Kunze et al.
 \cite{kunze2017transfer} answered this question by looking at the weight difference distribution per layer early and late into training.
-What they found was that models trained on a similar task to the target domain only had a minor change in the weight distribution.
-This reasoning could possibly be applied to more accurately find the optimal backbone.
+They found that models trained on a similar task to the target domain only had a minor change in the weight distribution.
+This reasoning could be applied to more accurately find the optimal backbone.
 Additionally, the use of a pre-trained model could be further explored by looking at the effect of the number of layers frozen during training.
 
+\newpage
 \section{Conclusion}
-To conclude, we noticed that first and foremost the task of predicting plant traits is a hard one.
-It requires careful consideration during every step of the process from understanding the data to how a neural network architecture will learn from it.
-From our results, we observed that a model trained on images and ancillary geodata converges faster but does not necessarily have a higher $R^2$ than a model trained on images alone.
-In the final test run these results were also observed with our image model having a higher $R^2$ than the combined model.
-However, both models still have a negative $R^2$ suggesting some core limitations in the models.
-Future work could look at the effect of different hyperparameters, model architectures, and data modifications to improve the prediction capabilities of the models.
+
+This paper aimed to evaluate the effectiveness of incorporating ancillary geodata alongside images in a Convolutional Neural Network (CNN) for predicting plant traits from images.
+Our key findings revealed that integrating both data types into models resulted in faster convergence, but it did not improve the predictive accuracy of models trained solely on images.
+This result indicates that adding geodata does not necessarily translate into better performance, and it highlights the complexity of the task at hand.
+
+We faced several challenges during our research, such as the negative R² values of the models, which suggest underfitting or other core limitations in our models.
+Additionally, varying image quality and the presence of outliers complicated model training and performance, highlighting the importance of data quality and preprocessing for developing effective predictive models.
+
+For future work, we suggest exploring alternative hyperparameters, diverse model architectures, and better data augmentation techniques.
+Incorporating depth estimation for scale normalization could be a promising method for addressing the variability present in image-based data.
+Investigating the adaptability of pre-trained models to this specific domain could also provide valuable insights into optimizing model training and performance.
+
+Our research highlights the potential of machine learning in ecological research, specifically in understanding and responding to the impact of climate change on plant life.
+Despite the challenges, continued innovation and exploration can uncover deeper insights into the planet's biological responses to a changing climate, which is crucial for global conservation and sustainability efforts.
 

### 2024-03-29 paper/info_sheet.tex
added info sheet

@@
 \hline
-Kai Niermann & 1234567 \\
-Dávid Miklo & 1234567 \\
-Trix Taiclet & 1234567 \\
-Conner Dassen & 1234567 \\
-Red Kaláb & 1234567 \\
+Kai Niermann & 2720905 \\
+Dávid Miklo & 2767001 \\
+Trix Taiclet & 2785891 \\
+Conner Dassen & 2695049 \\
+Red Kaláb & 2735630 \\
 \end{tabular}
@@
 
-\paragraph{Use of AI tools} \textit{If you used tools like ChatGPT or Github Copilot in any phase of the report, please detail here which tools you used, what parts of the work you used them for and how.
-Note that if you use an AI tool and don't report it, it's considered fraud, so be detailed.
-Note also that having an AI tool write for you is not allowed.
-}
+\paragraph{Use of AI tools} \textit{}
+\paragraph{Link to code (optional)} \textit{}
 
-\paragraph{Link to code (optional)} \textit{If you would like to share your code, you can link to your repository here.}
-
-\paragraph{Group disagreements} \textit{If there are any disagreements in the group, you may \textbf{not} remove a student from the author list without their consent.
-You should bring disagreements to our attention early.
-As a last resort, you can describe any grievances here}
+\paragraph{Group disagreements} \textit{}
